---
title: "ST 558 Project 1"
format: html
editor: visual
---

## Data Processing

```{r}
# Load packages
library(tidyverse)
library(readxl)
library(DBI)
```

```{r}
# Read in the data
data1 <- read_csv("EDU01a.csv")
```

#### 1. Only select the columns identified in the instructional document
```{r}
# Name the data frame
selected_data1 <- data1 |>

# Select the respective columns
  select(Area_name, STCOU, ends_with("D")) |>

# Rename the Area_name variable
  rename(area_name = Area_name)
```

#### 2. Convert the data into long format
```{r}
long_data1 <- pivot_longer(selected_data1, cols = ends_with("D"), names_to = "Survey", values_to = "Enrollment Value")
```

#### 3. Survey values
```{r}
long_data1 |>
  
  # Extract the year from the 'Survey' column, covert to a number, and create a new 'Year' variable with these values
  mutate(Year = as.numeric(substr(Survey, start = 8, stop = 9))) |>
  
  # Covert the two digit year into a four digit year and overwrite the 'Year' column
  mutate(Year = ifelse(Year > 24, Year + 1900, Year + 2000)) |>
  
  # Extract the first 7 digits from the 'Survey' column and create a new 'Measurement' variable with these values
  mutate(Measurement = substr(Survey, start = 1, stop = 7))
```

#### 4. Create two data sets

**Data set that contains only non-county data**
```{r}
# Identify the indices corresponding to states
county_indices <- grep(pattern = ", \\w\\w", long_data1$area_name)

# Create the county data set with only non-county level data
noncounty_tibble <- long_data1[-county_indices, ]

# Add a class to the tibble called 'state'
class(noncounty_tibble) <- c("state", class(noncounty_tibble))
```

**Data set that contains only county level data**
```{r}
# Create the county data set with only county level data
county_tibble <- long_data1[county_indices, ]

# Add a class to the tibble called 'county'
class(county_tibble) <- c("county", class(county_tibble))
```

#### 5. Create a new variable to describe the corresponding state in the county tibble
```{r}
county_tibble |>
  mutate(State = substr(area_name, start = nchar(area_name)-1, stop = nchar(area_name)))
```

#### 6. Create a new 'division' variable for the non-county tibble
```{r}
noncounty_tibble |>
  mutate(Division = ifelse(area_name %in% c("CONNECTICUT", "MAINE", "MASSACHUSETTS", "NEW HAMPSHIRE", "RHODE ISLAND", "VERMONT"), "Division 1",
                           ifelse(area_name %in% c("NEW JERSEY", "NEW YORK", "PENNSYLVANIA"), "Division 2",
                                  ifelse(area_name %in% c("ILLINOIS", "INDIANA", "MICHIGAN", "OHIO", "WISCONSIN"), "Division 3",
                                         ifelse(area_name %in% c("IOWA", "KANSAS", "MINNESOTA", "MISSOURI", "NEBRASKA", "NORTH DAKOTA", "SOUTH DAKOTA"), "Division 4",
                                                ifelse(area_name %in% c("DELAWARE", "FLORIDA", "GEORGIA", "MARYLAND", "NORTH CAROLINA", "SOUTH CAROLINA", "VIRGINIA", "DISTRICT OF COLUMBIA", "District of Columbia", "WEST VIRGINIA"), "Division 5",
                                                       ifelse(area_name %in% c("ALABAMA", "KENTUCKY", "MISSISSIPPI", "TENNESSEE"), "Division 6",
                                                              ifelse(area_name %in% c("ARKANSAS", "LOUISIANA", "OKLAHOMA", "TEXAS"), "Division 7",
                                                                     ifelse(area_name %in% c("ARIZONA", "COLORADO", "IDAHO", "MONTANA", "NEVADA", "NEW MEXICO", "UTAH", "WYOMING"), "Division 8",
                                                                            ifelse(area_name %in% c("ALASKA", "CALIFORNIA", "HAWAII", "OREGON", "WASHINGTON"), "Division 9", "ERROR"))))))))))
```

## Functions

#### 1. Function to select necessary columns and convert the data into long format
```{r}
long_format_conversion <- function(df, value = "Enrollment Value") {
  selected_data <- df |>

# Select the respective columns
  select(Area_name, STCOU, ends_with("D")) |>

# Rename the Area_name variable
  rename(area_name = Area_name)
  
# Convert the data into long format
  long_data <- pivot_longer(selected_data, cols = ends_with("D"), names_to = "Survey", values_to = value)
  return(long_data)
}
```

```{r}
# Function one test
data2 <- read_csv("EDU01B.csv")
long_data2 <- long_format_conversion(data2)
long_data2
```

#### 2. Function for pulling out year and measurement from the survey values
```{r}
survey_function <- function(long_df) {
  long_df |>
    
  # Extract the year from the 'Survey' column, covert to a number, and create a new 'Year' variable with these values
  mutate(Year = as.numeric(substr(Survey, start = 8, stop = 9))) |>
    
  # Covert the two digit year into a four digit year and overwrite the 'Year' column
  mutate(Year = ifelse(Year > 25, Year + 1900, Year + 2000)) |>
    
  # Extract the first 7 digits from the 'Survey' column and create a new 'Measurement' variable with these values
  mutate(Measurement = substr(Survey, start = 1, stop = 7))
}
```

```{r}
# Function two test
survey_function(long_data2)
```









